---
title: "Homework 02: Explore Gapminder and use dplyr"
author: Jessica Schaub
date: September 19, 2018
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Overview
This is an R Markdown file for Homework 02, with the purpose of exploring the `gapminder` dataset using `dplyr`and `ggplot`.

## Load Gapminder Data
```{r,include=FALSE}
library(gapminder)
library(tidyverse)
```
Using the library function, we can load our `gapminder` data series and the `tidyverse` package that contains `ggplot` and `dplyr`.

## Explore Gapminder
Here, we will examine the gapminder object using a variety of functions. First, let's explore using base functions in R:
```{r}
# What type of object
typeof(gapminder)
#class of object
class(gapminder)
```
Using the `class()` and `typeof()` functions, we learn that `gapminder` is a list, which is an object that contains different types of data (numerical, character, vectors, etc...)
We also learned that `gapminder` has three classes: tbl_df, tbl, and data.frame.

Now let's examine the data and format of `gapminder`:
```{r}
#glimpse shows useful info about object, dplyr function
glimpse(gapminder)
```

Using this one table, we can answer many questions about the data.

- There are 6 variable/columns
- There are 1704 observations/rows
- Country & continent are factors, year & population are integers, and life expectancy and gdp per capita are doubles

Of course, there are other functions that we can use to examine the data. Nrow and ncol can be used for a simple printout, and are useful if that is the only information you are looking for. 
```{r}
nrow(gapminder)
ncol(gapminder)
```

We can also print the data out, which gives us information on the data type of variables. 
We can also count rows and columns ourselves with this view, but it's difficult to imagine a scenario where that would be useful.

```{r}
print(gapminder)
```

Overall, the `dplyr` function `glimpse()` was most useful for examining the data.

## Explore Individual Variables

### Categorical Variables
Continent and Country are both categorical variables in `gapminder`.
Let's examine the number of possible values:
```{r}
#n_distinct gives the number of distinct values within a variable
n_distinct(gapminder$continent)
n_distinct(gapminder$country)
```

Since there are a large number of possible values for country, we will choose continent as our categorical variable.
The smaller number of values will make plotting cleaner and easier.
Let's view the specific values of continent:
```{r}
#Levels prints out every unique value for categorical variables
levels(gapminder$continent)
```

### Quantitative Variable

The other four variables are quantitative: year, population, GDP per capita, and life expectancy.
We can examine the range of each variable:

```{r}
#min and max values for year
paste("Year: min = ",min(gapminder$year),", max = ",max(gapminder$year))

#min and max values for population
paste("Population: min = ",min(gapminder$pop),", max = ",max(gapminder$pop))

#min and max values for GDP per capita
paste("GDP per Capita: min = ",min(gapminder$gdpPercap),", max = ",max(gapminder$gdpPercap))

#min and max values for life expectancy
paste("Life Expectancy: min = ",min(gapminder$lifeExp),", max = ",max(gapminder$lifeExp))
```

Life expectancy has a nice range that can also provide informative results, so we can use this as our numerical variable.
Let's examine Life Expectancy a bit closer:
```{r}
#plotting life expectancy as a histogram to examine spread and distribution
ggplot(data=gapminder, aes(x=lifeExp)) +
  geom_histogram(binwidth = 3, fill="black", col="grey") +
  geom_vline(aes(xintercept=mean(gapminder$lifeExp)), color="red", linetype="dashed") +
  ylab("Count") + 
  xlab("Life Expectancy (years)") + 
  ggtitle("Histogram of Life Expectancy") +
  theme_classic()

#red dashed line for mean
```
This histogram shows the distribution of the variable, and the mean is indicated by the dashed red line.
For more information on the statistics of the variable, we can print a summary:
```{r}
summary(gapminder$lifeExp)
```


## Explore Plot Types

### Two numerical variables

In this section, we will explore the relationship between year and life expectancy to see how life expectancy has changed with time.

```{r}
ggplot(data=gapminder, aes(x=year,y=lifeExp)) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE, color='red') +
  xlab("Year") + ylab("Life Expectancy (years)") + ggtitle("Life Expectancy over Time") +
  theme_classic()
```

This plot shows that life expectancy has increased with time across the whole world.
Now let's look at the trends in each continent to see if there are new patterns.

```{r}
ggplot(data=gapminder, aes(x=year,y=lifeExp, fill=continent, color=continent)) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE) +
  xlab("Year") + ylab("Life Expectancy (years)") + ggtitle("Life Expectancy over Time") +
  theme_classic()
  
```

From this graph, we can gather more information. For example, Africa has consistently shown the lowest life expectancy and had the slowest increase over time.
Asia has the fastest increase in life expectancy since 1951.
Meanwhile, Oceania has consistently shown the highest life expectancy.

### One numerical and one categorical variable

In this section, we will explore the relationship between life expectancy and continent for the year 2007.

```{r}
#obtain values for 2007
gapminder %>% 
  filter(year==2007) %>% #pipe into ggplot
  ggplot(aes(x=continent, y=lifeExp)) +
  geom_boxplot(varwidth = TRUE) +
  xlab("Continent") + ylab("Life Expectancy (years)") +
  ggtitle("Average Life Expectancy for Each Continent in 2007") +
  theme_classic()
```

This box plot gives important statistical information, including mean and quantiles.
The width of the box also represents the relative number of observations for each continent.

## Practice with Piping, Filtering, and Selecting

Some of `dplyr`'s most useful functions include `filter()`, `select()`, and the ability to pipe with `%>%`. 

We can filter for specific rows:
```{r}
#filter for something simple, like Canada
filter(gapminder, country=="Canada")

#filter for all Oceania and Africa data before 1977
filter(gapminder, continent %in% c('Oceania','Africa'), year < 1977)
```

We can also select different columns:
```{r}
#select for only gdpPercap and continent
select(gapminder, gdpPercap, continent)

#or we can select only columns that contain the letter E
select(gapminder,contains("e"))
```

Piping also allows us to combine these functions and subset our data very specifically and efficiently:

```{r}
#Find all values for Zimbabwe between 1975 and 2000, and only plot variables that contain the letter 'r'
gapminder %>% 
  filter(country == 'Zimbabwe', year %in% c(1975:2000)) %>% 
  select(contains("r"))
```





